---
title: "Spearman's ranking correlation"
author: "Fernando Náufel"
date: now
date-format: "DD/MM/YYYY HH:mm"
lang: en

bibliography: bibliography.bib
csl: chicago-fullnote-bibliography.csl

format:
  html:
    toc: true  
    toc-depth: 3
    number-depth: 3
    theme: 
      - journal
      - _custom.scss
    link-external-icon: true
    link-external-newwindow: true
    link-external-filter: ^(?:http:|https:)\/\/fnaufel\.github\.io
    df-print: paged
    code-link: true
    code-copy: true
    code-tools: true
    self-contained: true
---

{{< include _math.qmd >}}

# Setup

```{r setup}
source('_setup.R')
```


# Definition

Spearman's $\rho$ between random variables $X$ and $Y$ is simply the linear correlation of the associated rank variables $R(X)$ and $R(Y)$.


# Functions

```{r}
example <- function(x, y) {
  
  cat('x          =', x, '\n')
  cat('y          =', y, '\n')
  cat('linear cor =', cor(x, y), '\n\n')

  cat('rank(x)    =', rank(x), '\n')
  cat('rank(y)    =', rank(y), '\n')
  cat('spear. cor =', cor(x, y, method = 'spearman'), '\n\n')
    
}

scatter_plot <- function(x, y) {
  
  tibble(x, y) %>% 
    ggplot(aes(x, y)) +
      geom_point() +
      geom_smooth(formula = y ~ x, method = 'lm', se = FALSE) +
      labs(title = 'Original values')
  
}

scatter_plot_r <- function(x, y) {
  
  x <- rank(x)
  y <- rank(y)
  
  tibble(x, y) %>% 
    ggplot(aes(x, y)) +
      geom_point() +
      geom_smooth(formula = y ~ x, method = 'lm', se = FALSE) +
      labs(title = 'Ranks')
  
}

```


# Examples

## Very small

```{r}
x <- c(1, 2, 3)
y <- c(20, 25, 5)
example(x, y)
```

```{r}
scatter_plot(x, y)
scatter_plot_r(x, y)
```


## With ties

```{r}
x <- c(1, 2, 2, 3)
y <- c(20, 21, 5, 6)
example(x, y)
```

Ties are not a problem.

```{r}
scatter_plot(x, y)
scatter_plot_r(x, y)
```

```{r}
y <- c(1, 2, 2, 3)
x <- c(20, 21, 5, 6)
example(x, y)
```

```{r}
scatter_plot(x, y)
scatter_plot_r(x, y)
```


## People ranked two ways

```{r}
set.seed(1234)

ranking_1 <- sample(LETTERS)
ranking_2 <- sample(LETTERS)

cat('Ranking 1:', ranking_1, '\n')
cat('Ranking 2:', ranking_2, '\n')
```

I need to write these rankings as pairs:

* Person `A`: $(`r match('A', ranking_1)`, `r match('A', ranking_2)`)$
* Person `B`: $(`r match('B', ranking_1)`, `r match('B', ranking_2)`)$
* etc.

This can be done by the `order` function:

```{r}
order(ranking_1)
order(ranking_2)
```

```{r}
pairing <- tibble(
  person = LETTERS,
  r1 = order(ranking_1),
  r2 = order(ranking_2)
)

pairing
```

```{r}
scatter_plot(pairing$r1, pairing$r2)
```

The plot says "original values", but these are already the ranks!

This is why the scatter plot of the ranks looks exactly the same:

```{r}
scatter_plot_r(pairing$r1, pairing$r2)
```

```{r}
example(pairing$r1, pairing$r2)
```

::: {.callout-note title="When `rank` is idempotent"}

If $x$ contains all integer values between $1$ and $n$, with no duplicates, then $\operatorname{rank}(x) = x$.

:::


## An example similar to DWARFF

See <https://fnaufel.github.io/2024-DWARFF-Romulo/correlacao-ranking-expert-list.html>.

???

